Three distinct patterns of mental health response following accidents in mountain sports – a follow-up study of individuals treated at a tertiary trauma center

### 2023-11-07 paper/markdown/supplementary_material.Rmd
Feedback, October 2023, markdown and figures

For semi-supervised clustering, diagnostic and performance of the clustering analysis, the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) with implementation of algorithms and distance measures from _factoextra_, _cluster_ and _philentropy_ [@Schubert2019; @Drost2018; @Kassambara2020] was utilized. 
UMAP (uniform manifold approximation and projection) was done with the _umap_ and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) packages [@Konopka2022; @McInnes2018]. 
Cross-distances between the training and test subsets of the study cohort and cross-distances between clusters in the training and test subsets of the study cohort were computed with _philentropy_ [@Drost2018] and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools).

@@

The psychometric battery (__Supplementary Table S\@ref(tab:tab-mental-battery)__) consisted of German versions of assessment tools for 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ: patient health questionnaire, PHQ-9) [@Lowe2002; @Grafe2004], 
panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
persistent somatic symptoms (PHQ-15) [@Kroenke2002], 
resilience (RS13: 13-item resilience scale) [@Leppert2008], 
loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
post-traumatic stress disorder (PCL-5 DSM-5: PTSD checklist for DSM-5) [@Bovin2016].

Clinically relevant symptoms of anxiety were defined as GAD-7 $geq$ 10, 
clinically relevant symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], 
significant persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002]. 
Resilience classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008].

Items of the EUROHIS QOL 8 tool were scored as with 1 - 5 likert scales (1: extreme concerns, no satisfaction at all, 5: no concerns/full satisfaction) with each item representing a single domain of quality of life (quality of life, health, energy, finances, activity, self-esteem, relationship and housing). 
The total EUROHIS QOL score was defined as the arithmetic mean of all items [@Schmidt2006].

@@

Separate scores were calculated for domains B, C, D and E of the PCL-5 DSM-5 tool along with the total score being the sum of all items. 
Each PCL-5 DSM-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
Participants positive for the B domain or C domain PTSD symptoms were identified by at least one item per domain scored with 'moderate' or higher. 
Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 DSM-5 domains [@Bovin2016].

Traumatic events prior to the sport accident were assessed with the DIA-X tool (Diagnostic Expert System) [@Maercker2005]. 
Direct personal experience or being a witness of a traumatic event specified by the DIA-X questionnaire or by an additional yes/no item ('other traumatic events') was scored as 1. 
Prior traumatic event was assumed with at least one DIA-X item scored with 1. 
Prior sport accidents, 
flashbacks during mountain sport activity, 
confusion during mountain sport activity, 
self-reported need for psychological support, 
psychological/psychiatric support/therapy after the accident 
and presence of persistent physical health consequences after the accident 
were surveyed as single yes/no items. 
Flashbacks frequency during mountain sport activity was assessed in the following categories: none, more than one per year and more than one per month. 
Smoking was surveyed as a single yes/no question. 
Alcohol use was investigated with the CAGE tool with $\geq$ 2 points indicative of problematic alcohol consumption [@OBrien2008]. 
Data on the type of the accident date and daytime, accident mountain sport, injury diagnosis, injured body regions, injury severity, hospital treatment, surgery and number of ICD-10 surgical diagnoses  were extracted from electronic patient's records. 
Injury severity was assessed with the abbreviated injury scale (AIS) [@Gennarelli2006].

@@

* Cohen's $\kappa$ reliability: none: < 0.2, minimal: 0.2 - 0.4, weak: 0.4 - 0.6, moderate: 0.6 - 0.8, strong: $\geq$ 0.8

@@

Observations of the training subset of the study cohort were subjected to clustering in respect to the numeric psychometric scores (__Supplementary Table S\@ref(tab:tab-mental-battery)__). 
The score values were normalized and median-centered (function `center_data()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
For clustering, the PAM (partition around medoids) algorithm [@Schubert2019] with the cosine distance measure between observations [@Drost2018] was employed. 
The clustering object was constructed with the function `kcluster()` from the  [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
The choice of the clustering algorithm was motivated by its good explanatory performance measured by the fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares, method `var()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) and superior reproducibility in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier, method `cv()`, [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__. 
The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) [@Rousseeuw1987; @Kassambara2020]. 
By this means, three mental clusters were defined: the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) cluster (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__).

Assignment of the training subset observations to the mental clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
This semi-supervised procedure yielded clustering structures with similar fractions of explained variance and similar average silhouette statistic [@Rousseeuw1987] in the 
training (variance = `r signif(semi_clust$stats$variance[1], 2)`, 
average silhouette = `r signif(semi_clust$stats$mean_sil[1], 2)`) and 
test subset (variance = `r signif(semi_clust$stats$variance[2], 2)`, 
average silhouette = `r signif(semi_clust$stats$mean_sil[2], 2)`). 
Comparably good separation of the clusters could be discerned in the training and test subsets by 
a visual analysis of UMAP layouts and 
pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__). 
Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)__). 
An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental clusters (i.e. neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g. neutral vs PTG) (distances computed with the function `cross_distance()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools); __Supplementary Figure S\@ref(fig:fig-cosine)B__). 
Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).

Differences in frequencies of mental disorder symptoms as well as demographic, socioeconomic, clinical, accident- and recovery-related factors between the mental clusters were assessed in the entire cohort by $\chi^2$ test with Cramer V effect size statistic and by Kruskal-Wallis test with $\eta^2$ effect size statistic for categorical and numeric variables, respectively (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Supplementary Table S\@ref(tab:tab-clust-bcg)__).

@@

Two types of multi-parameter machine learning classifiers of the mental cluster assignment were developed in the training subset:

1. models employing candidate early predictors of cluster assignment, i.e. demographic, socioeconomic, medical history and accident-related explanatory factors available during acute medical management of the accident

@@

Of note, psychometric variables used for definition of the mental clusters as well as mental disorder symptoms, presence and frequency of flashbacks were excluded from the explanatory variable sets. 
The explanatory variables are listed in __Supplementary Table S\@ref(tab:tab-mod-variables)__.

@@

```{r tab-analysis-vars, tab.cap = 'Variables used in the analysis pipeline.'}

@@

flextable::flextable(suppl_tables$mental_battery) %>% 
  set_widths(c(2.3, 7, 10)) %>% 
  merge_v(1) %>% 
  footnote(1, 2:3, 
           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(suppl_tables$consistency) %>% 
  set_widths(c(3.5, 3, 3)) %>% 
   footnote(1, 1, 
           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5 DSM-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(suppl_tables$no_response) %>% 
  set_widths(c(3.2, 5.8, 5.8, 2.5, 2.2)) %>% 
   footnote(1, 1, 
           value = as_paragraph('AIS: abbreviated injury scale'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. P values corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-clust-fct, tab.cap = 'Differences in psychometric clustering factors between the mental clusters. Numeric variables are presented as medians with interquartile ranges (IQR). Statistical significance was determined by false discovery rate-corrected Kruskal-Wallis test with eta-square effect size statistic. The table is available in a supplementary Excel file.'}

@@

```{r tab-clust-bcg, tab.cap = 'Significant differences in demographic, socioeconomic, clinical, accident- and recovery-related factors, and mental disorder symptoms between the mental clusters in the entire cohort. Numeric variables are presented as medians with interquartile ranges (IQR). Categorical variables are presented as percentages and counts within the complete observation set.'}

flextable::flextable(suppl_tables$clust_bcg) %>% 
  set_widths(c(3, rep(4.3, 3), 2.5, 2.3)) %>% 
  footnote(1, 1, 
           value = as_paragraph('AIS: abbreviated injury score; PTSD: post-traumatic stress disorder; RS13: 13-item resilience scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 5:6, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. P values were corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-mod-variables, tab.cap = 'Sets of explanatory factors used for modeling of the mental cluster assignment.'}

flextable::flextable(suppl_tables$mod_variables) %>% 
  set_widths(c(2.5, 15.5)) %>% 
  footnote(1, 1, 
           value = as_paragraph('early predictors: variables available during acute medical management of the accident; full predictor set: variables available during acute medical management of the accident and during follow-up.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(suppl_tables$tuning) %>% 
  set_widths(c(2, 2.5, 8.5)) %>% 
  merge_v(1) %>%
  footnote(1, 1, 
           value = as_paragraph('early predictors: variables available during acute medical management of the accident; full predictor set: variables available during acute medical management of the accident and during follow-up.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('RF: random forest; NNet: neural network with a single hidden layer; SVM/radial: support vector machines with radial kernel; RPart: recursive partitioning; SDA: shrinkage discriminant analysis; cForest: conditional random forest; Elastic Net: elastic net multinomial regression.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-early-classifiers, tab.cap = 'Performance statistics of machine learning classifiers at predicting the mental cluster assignment. Models employing early predictors available during acute medical management of the accident.'}

@@

```{r tab-full-classifiers, tab.cap = 'Performance statistics of machine learning classifiers at predicting the mental cluster assignment. Models employing the full predictor set available during acute medical management of the accident and follow-up.'}

@@

```{r fig-clust-dev, fig.width = figur::convert(suppl_figures$clust_dev, to = 'in')$w, fig.height = figur::convert(suppl_figures$clust_dev, to = 'in')$h, fig.cap = 'Definition of the mental clusters in the training subset of the study cohort.'}

@@

__Supplementary Figure S\@ref(fig:fig-clust-dev). Definition of the mental clusters in the training subset of the study cohort.__

_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations._

@@

_The mental clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations. Assignment of the test subset observations to the mental clusters was done with the inverse distance weighted 27-nearest neighbors classifier. By this procedure, three mental clusters were identified: neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress). Numbers of observations in the mental clusters are shown in the plot legend in (A)._

@@

_(B) Pairwise cosine distances between observations in the mental clusters of the training and test subsets presented in heat maps._

@@

```{r fig-cosine, fig.width = figur::convert(suppl_figures$cosine, to = 'in')$w, fig.height = figur::convert(suppl_figures$cosine, to = 'in')$h, fig.cap = 'Distribution of the mental clusters and cosine distances between the mental clusters in the training and test subset of the study cohort.'}

@@

__Supplementary Figure S\@ref(fig:fig-cosine). Distribution of the mental clusters and cosine distances between the mental clusters in the training and test subset of the study cohort.__

_Study cohorts observations were assigned to the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) clusters by semi-supervised clustering as presented in Supplementary Figure S\@ref(fig:fig-semi-clust)._

_(A) Percentages of observations in the training and test subsets of the study cohort in the mental clusters. Numbers of complete observations are indicated in the Y axis._

_(B) Cosine cross-distances between the mental clusters in the training and test subset of the study dataset visualized as a heat map. Cross-distances are indicated in the tiles._

@@

```{r fig-feat-hm, fig.width = figur::convert(suppl_figures$feat_hm, to = 'in')$w, fig.height = figur::convert(suppl_figures$feat_hm, to = 'in')$h, fig.cap = 'Levels of psychometric scores used for the cluster definition in the mental clusters.'}

@@

__Supplementary Figure S\@ref(fig:fig-feat-hm). Levels of psychometric scores used for the cluster definition in the mental clusters.__

>>>>

For semi-supervised clustering, diagnostic and performance testing of the clustering analysis, the package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) with implementation of algorithms and distance measures from _factoextra_, _cluster_ and _philentropy_ [@Schubert2019; @Drost2018; @Kassambara2020] was utilized. 
UMAP (uniform manifold approximation and projection) was done with the _umap_ and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) packages [@Konopka2022; @McInnes2018]. 
Cross-distances between the training and test subsets of the study cohort and cross-distances between clusters in the training and test subsets of the study cohort were computed with _philentropy_ [@Drost2018] and [_clustTools_](https://github.com/PiotrTymoszuk/clustTools).

@@

The psychometric battery (__Supplementary Table S\@ref(tab:tab-mental-battery)__) consisted of German versions of assessment tools for 
anxiety (GAD-7: 7-item general anxiety disorder scale) [@Spitzer2006], 
depression (PHQ: patient health questionnaire, PHQ-9) [@Lowe2002; @Grafe2004], 
panic (PHQ-panic module) [@Lowe2002; @Grafe2004], 
common persistent somatic symptoms as a substitute for somatization (PHQ-15) [@Kroenke2002], 
resilience (RS13: 13-item resilience scale) [@Leppert2008], 
loss of sense of coherence (SOC-9L: Leipzig 9-item sense of coherence questionnaire) [@Schumacher2000], 
quality of life (EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale) [@Schmidt2006], 
post-traumatic growth (PTGI: post-traumatic growth inventory) [@Tedeschi1996] and 
post-traumatic stress disorder (PCL-5: PTSD checklist for DSM-5) [@Bovin2016].

Clinically relevant symptoms of anxiety were defined as GAD-7 $\geq$ 10, 
clinically relevant symptoms of depression were defined as PHQ-9 $\geq$ 11 [@Manea2012], 
significant persistent somatic symptoms were defined as PHQ-15 $\geq$ 11 [@Kroenke2002]. 
Resilience classes were defined as follows: low: RS-13 0 - 65, moderate: 66 - 72, high: $\geq$ 73 [@Leppert2008].

Items of the EUROHIS QOL 8 tool were scored as with 1 - 5 Likert scales (1: extreme concerns, no satisfaction at all, 5: no concerns/full satisfaction) with each item representing a single domain of quality of life (quality of life, health, energy, finances, activity, self-esteem, relationship and housing). 
The total EUROHIS QOL score was defined as the arithmetic mean of all items [@Schmidt2006].

@@

Separate scores were calculated for domains B, C, D and E of the PCL-5 tool along with the total score being the sum of all items. 
Each PCL-5 item was scored as 0: not at all, 1: a little bit, 2: moderate, 3: quite a bit and 4: extremely. 
Participants positive for the B domain or C domain PTSD symptoms were identified by at least one item per domain scored with 'moderate' or higher. 
Participants positive for the D or E domain PTSD symptoms were identified by at least two items per domain scored with 'moderate' or higher. 
Significant PTSD symptoms where assumed in participants screened positive for at least one of the B, C, D or E PCL-5 domains [@Bovin2016].

Traumatic events prior to the mountain sport accident were assessed with the DIA-X tool (Diagnostic Expert System) [@Maercker2005]. 
Direct personal experience or being a witness of a traumatic event specified by the DIA-X questionnaire or by an additional yes/no item ('other traumatic events') was scored as 1. 
Prior traumatic event was assumed with at least one DIA-X item scored with 1. 
Pre-existing mental disorder diagnosed by a physician, 
prior mountain sport accidents, 
flashbacks during mountain sport activity, 
confusion during mountain sport activity, 
self-reported need for psychological support following the accident, 
psychological/psychiatric support/therapy after the accident 
and presence of persistent physical health consequences related to the accident 
were surveyed as single yes/no items. 
Flashbacks frequency during mountain sport activity was assessed in the following categories: none, more than one per year and more than one per month. 
Smoking was surveyed as a single yes/no question. 
Alcohol use was investigated with the CAGE tool with $\geq$ 2 points indicative of problematic alcohol consumption [@OBrien2008]. 
Data on the type of the accident date and daytime, accident mountain sport, injury diagnosis, injured body regions, injury severity, hospital treatment, surgery and number of ICD-10 surgical diagnoses  were extracted from electronic patient's records. 
Injury severity was assessed with the abbreviated injury scale (AIS) [@Gennarelli2006].

@@

* Cohen's $\kappa$ inter-rater reliability statistic: none: < 0.2, minimal: 0.2 - 0.4, weak: 0.4 - 0.6, moderate: 0.6 - 0.8, strong: $\geq$ 0.8

@@

Observations of the training subset of the study cohort were subjected to clustering in respect to the numeric psychometric scores (__Supplementary Table S\@ref(tab:tab-mental-battery)__). 
The score values were normalized and median-centered (function `center_data()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools)). 
For clustering, the PAM (partition around medoids) algorithm [@Schubert2019] with the cosine distance measure between observations [@Drost2018] was employed. 
The clustering object was constructed with the function `kcluster()` from the  [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package.
The choice of the clustering algorithm was motivated by 
its good explanatory performance measured by the fraction of explained clustering variance (ratio of between-cluster sum of squares to total sum of squares), 
good separability between the clusters measured by mean silhouette width [@Rousseeuw1987], 
and superior reproducibility in 10-fold cross-validation [@Lange2004] (cluster assignment in the folds by an inverse distance weighted 27-nearest neighbors classifier) in a comparison with several other clustering algorithms presented in __Supplementary Figure S\@ref(fig:fig-clust-dev)A__. 
The explained variance, mean silhouette width and cross-validated accuracy statistics were computed with the methods `var()`, `silhouette()` and `cv()` from the [_clustTools_](https://github.com/PiotrTymoszuk/clustTools) package. 
The number of cluster was chosen based on the bend of the curve of within-cluster sum of squares and the peak of mean silhouette statistic (method `plot()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA)) [@Rousseeuw1987; @Kassambara2020]. 
By this means, three mental health clusters were defined: the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) cluster (__Supplementary Figure S\@ref(fig:fig-clust-dev)B__).

Assignment of the training subset observations to the mental health clusters was accomplished with an inverse distance weighted 27-nearest neighbor classifier. 
This semi-supervised procedure yielded clustering structures with similar fractions of explained variance and similar average silhouette statistic [@Rousseeuw1987] in the 
training (variance = `r signif(semi_clust$stats$variance[1], 2)`, 
average silhouette = `r signif(semi_clust$stats$mean_sil[1], 2)`) and 
test subset (variance = `r signif(semi_clust$stats$variance[2], 2)`, 
average silhouette = `r signif(semi_clust$stats$mean_sil[2], 2)`). 
Comparably good separation of the clusters could be discerned in the training and test subsets by 
a visual analysis of UMAP layouts and 
pairwise distance heat maps (__Supplementary Figure S\@ref(fig:fig-semi-clust)__). 
Distribution of the cluster frequency was similar in the training and test subset (__Supplementary Figure S\@ref(fig:fig-cosine)__). 
An analysis of cosine cross-distances between the clusters in the training and test subsets revealed far higher similarity of the corresponding mental health clusters (i.e. neutral vs neutral, PTG vs PTG, PTS vs PTS) as compared with similarity of non-analogous clusters (e.g. neutral vs PTG) (distances computed with the function `cross_distance()`, package [_clustTools_](https://github.com/PiotrTymoszuk/clustTools); __Supplementary Figure S\@ref(fig:fig-cosine)B__). 
Finally, quality of semi-supervised clustering was investigated by comparison of normalized levels of the clustering variables between the mental health clusters in the training and test subset by Kruskal-Wallis test with $\eta^2$ effect size statistic (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Figure 2__, __Supplementary Figure S\@ref(fig:fig-feat-hm)__, __Supplementary Table S\@ref(tab:tab-clust-fct)__).

Differences in frequencies of mental disorder symptoms as well as demographic, socioeconomic, clinical, accident- and recovery-related factors between the mental health clusters were assessed in the entire cohort by $\chi^2$ test with Cramer V effect size statistic and by Kruskal-Wallis test with $\eta^2$ effect size statistic for categorical and numeric variables, respectively (function `compare_variables()`, package [_ExDA_](https://github.com/PiotrTymoszuk/ExDA); __Supplementary Table S\@ref(tab:tab-clust-symptoms)__ and __S\@ref(tab:tab-clust-bcg)__).

@@

Two types of multi-parameter machine learning classifiers of the mental health cluster assignment were developed in the training subset:

1. models employing candidate early predictors of cluster assignment, i.e. demographic, socioeconomic, medical history and accident-related explanatory factors available during acute medical management of the accident victim

@@

Of note, psychometric variables used for definition of the mental health clusters as well as symptoms of mental disorders, presence and frequency of flashbacks were excluded from the explanatory variable sets. 
The explanatory variables are listed in __Supplementary Table S\@ref(tab:tab-mod-variables)__.

@@

```{r tab-analysis-vars, tab.cap = 'Variables used in the analysis pipeline. The table is available as a supplementary Excel file.'}

@@

flextable::flextable(suppl_tables$mental_battery) %>% 
  set_widths(c(2.3, 7, 10)) %>% 
  merge_v(1) %>% 
  footnote(1, 2:3, 
           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ-9: 9-item patient health questionnaire for depressive symptoms; PHQ-15: 15-item patient health questionnaire for common somatic symptoms as a substitute for somatization; PHQ-panic: patient health questionnaire module for panic disorders; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(suppl_tables$consistency) %>% 
  set_widths(c(4.5, 3, 3)) %>% 
   footnote(1, 1, 
           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ-9: 9-item patient health questionnaire for depressive symptoms; PHQ-15: 15-item patient health questionnaire for common somatic symptoms as a substitute for somatization; PHQ-panic: patient health questionnaire module for panic disorders; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(suppl_tables$no_response) %>% 
  set_widths(c(3.2, 5.8, 5.8, 2.5, 2.2)) %>% 
   footnote(1, 1, 
           value = as_paragraph('AIS: abbreviated injury scale; ICD-10: 10th version of the International Statistical Classification of Diseases and Related Health Problems.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 4:5, 
           value = as_paragraph('Numeric variables: Mann-Whitney test with r effect size statistic; categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. P values corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-clust-fct, tab.cap = 'Differences in psychometric clustering factors between the mental health clusters. Numeric variables are presented as medians with interquartile ranges (IQR). Statistical significance was determined by false discovery rate-corrected Kruskal-Wallis test with eta-square effect size statistic. The table is available in a supplementary Excel file.'}

@@

```{r tab-clust-symptoms, tab.cap = 'Frequency of mental disorder symptoms in the mental health clusters in the entire cohort. Categorical variables are presented as percentages and counts within the clusters.'}

flextable::flextable(suppl_tables$clust_symptoms) %>% 
  set_widths(c(4, rep(4.3, 3), 2.5, 2.3)) %>% 
  footnote(1, 1, 
           value = as_paragraph('GAD-7: 7-item general anxiety disorder scale; PHQ-9: 9-item patient health questionnaire for depressive symptoms; PHQ-15: 15-item patient health questionnaire for common somatic symptoms as a substitute for somatization; PHQ-panic: patient health questionnaire module for panic disorders; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTSD: post-traumatic stress disorder.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 5:6, 
           value = as_paragraph('\u03C7\u00B2 test with Cramer V effect size statistic. P values were corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-clust-bcg, tab.cap = 'Differences in demographic, socioeconomic, clinical, accident- and recovery-related factors, and between the mental clusters in the entire cohort. Significant effects are presented, the full table is available as a supplementary Excel file. Numeric variables are presented as medians with interquartile ranges (IQR). Categorical variables are presented as percentages and counts within the clusters.'}

suppl_tables$clust_bcg %>%
  filter(!stri_detect(Significance, regex = '^ns')) %>% 
  flextable %>% 
  set_widths(c(3, rep(4.3, 3), 2.5, 2.3)) %>% 
  footnote(1, 1, 
           value = as_paragraph('psychological support need: subjective need for psychological or psychiatric support following the accident; physical health consequences: physical health consequences related to the accident.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 5:6, 
           value = as_paragraph('Numeric variables: Kruskal-Wallis test with \u03B7\u00B2 effect size statistic. Categorical variables: \u03C7\u00B2 test with Cramer V effect size statistic. P values were corrected for multiple testing with the false discovery rate method.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

```{r tab-mod-variables, tab.cap = 'Sets of explanatory factors used for modeling of the mental health cluster assignment.'}

flextable::flextable(suppl_tables$mod_variables) %>% 
  set_widths(c(2.5, 15.5)) %>% 
  footnote(1, 1, 
           value = as_paragraph('early predictors: variables available during acute medical management of the patient; full predictor set: variables available during acute medical management of the patient and during follow-up.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  theme_vanilla

@@

flextable::flextable(suppl_tables$tuning) %>% 
  set_widths(c(2, 2.5, 8.5)) %>% 
  merge_v(1) %>%
  footnote(1, 1, 
           value = as_paragraph('early predictors: variables available during acute medical management of the patient; full predictor set: variables available during acute medical management of the patient and during follow-up.'), 
           part = 'header', 
           ref_symbols = 'a') %>% 
  footnote(1, 2, 
           value = as_paragraph('RF: random forest; NNet: neural network with a single hidden layer; SVM/radial: support vector machines with radial kernel; RPart: recursive partitioning; SDA: shrinkage discriminant analysis; cForest: conditional random forest; Elastic Net: elastic net multinomial regression.'), 
           part = 'header', 
           ref_symbols = 'b') %>% 
  theme_vanilla

@@

```{r tab-early-classifiers, tab.cap = 'Performance statistics of machine learning classifiers at predicting the mental health cluster assignment. Models employing early predictors available during acute medical management of the patient.'}

@@

```{r tab-full-classifiers, tab.cap = 'Performance statistics of machine learning classifiers at predicting the mental health cluster assignment. Models employing the full predictor set available during acute medical management of the patient and follow-up.'}

@@

```{r fig-clust-dev, fig.width = figur::convert(suppl_figures$clust_dev, to = 'in')$w, fig.height = figur::convert(suppl_figures$clust_dev, to = 'in')$h, fig.cap = 'Definition of the mental health clusters in the training subset of the study cohort.'}

@@

__Supplementary Figure S\@ref(fig:fig-clust-dev). Definition of the mental health clusters in the training subset of the study cohort.__

_The mental health clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations._

@@

_The mental health clusters were defined in respect to psychometric scoring in the training subset of the study cohort by PAM (partition around medoids) with cosine distance between the observations. Assignment of the test subset observations to the mental health clusters was done with the inverse distance weighted 27-nearest neighbors classifier. By this procedure, three mental health clusters were identified: neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress). Numbers of observations in the mental health clusters are shown in the plot legend in (A)._

@@

_(B) Pairwise cosine distances between observations in the mental health clusters of the training and test subsets presented in heat maps._

@@

```{r fig-cosine, fig.width = figur::convert(suppl_figures$cosine, to = 'in')$w, fig.height = figur::convert(suppl_figures$cosine, to = 'in')$h, fig.cap = 'Distribution of the mental health clusters and cosine distances between the mental health clusters in the training and test subset of the study cohort.'}

@@

__Supplementary Figure S\@ref(fig:fig-cosine). Distribution of the mental health clusters and cosine distances between the mental health clusters in the training and test subset of the study cohort.__

_Study cohorts observations were assigned to the neutral, PTG (post-traumatic growth) and PTS (post-traumatic stress) mental health clusters by semi-supervised clustering as presented in Supplementary Figure S\@ref(fig:fig-semi-clust)._

_(A) Percentages of observations in the training and test subsets of the study cohort in the mental health clusters. Numbers of complete observations are indicated in the Y axis._

_(B) Cosine cross-distances between the mental health clusters in the training and test subsets of the study dataset visualized as a heat map. Average cross-distances with 95% percentile range are indicated in the tiles._

@@

```{r fig-feat-hm, fig.width = figur::convert(suppl_figures$feat_hm, to = 'in')$w, fig.height = figur::convert(suppl_figures$feat_hm, to = 'in')$h, fig.cap = 'Levels of psychometric scores used for the cluster definition in the mental health clusters.'}

@@

__Supplementary Figure S\@ref(fig:fig-feat-hm). Levels of psychometric scores used for the cluster definition in the mental health clusters.__

_Levels of psychometric scores used in clustering in the mental health clusters. Normalized, median-centered score values (Z scores) are presented as heat maps. Numbers of observations in the clusters are displayed in the plot captions. PSS4: 4-item perceived stress scale; GAD-7: 7-item general anxiety disorder scale; PHQ: patient health questionnaire; EUROHIS-QOL 8: 8-item EUROHIS project quality of life scale; SOC-9L: Leipzig 9-item sense of coherence questionnaire; RS13: 13-item resilience scale; PCL-5: PTSD checklist for DSM-5; PTGI: post-traumatic growth inventory; PTSD: post-traumatic stress disorder._
